---
title: "General Data for Model"
author: "Erika Nares"
date: "2026-05-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(reticulate)
```

1. Getting game log data 
```{python}
from nba_api.stats.endpoints import leaguegamelog

def get_game_logs(season_year):

    logs = leaguegamelog.LeagueGameLog(
        season = season_year,
        season_type_all_star = "Regular Season",
        player_or_team_abbreviation = "T",
        timeout = 60
    )

    df_logs = logs.get_data_frames()[0]

    return df_logs
```

2. Getting team data 
```{python}
from nba_api.stats.endpoints import leaguedashteamstats

def get_teams(season_year):

    teams = leaguedashteamstats.LeagueDashTeamStats(
        season = season_year,
        season_type_all_star = "Regular Season"
    )

    df_teams = teams.get_data_frames()[0]

    return df_teams
```


3. Retreive SOS (Strength of Schedule) to calculate SRS (Simple Rating System = Average Point Differential + SOS)
```{r}
library(dplyr)
library(stringr)

get_sos <- function(season_year){
  
  df_logs <- py$get_game_logs(season_year)
  df_teams <- py$get_teams(season_year)
  
  # Make lookup table to find the full name from abbreviations
  abbrev_lookup <- df_logs %>%
    select(TEAM_ABBREVIATION, TEAM_NAME) %>%
    distinct() %>%
    rename(
      OPP_ABBREVIATION = TEAM_ABBREVIATION,
      OPP_TEAM_NAME = TEAM_NAME
    )
  
  # Get opponent abbreviation from matchup in df_logs 
  df_logs <- df_logs %>%
    select(
      TEAM_NAME,
      TEAM_ABBREVIATION,
      MATCHUP
    ) %>%
    mutate(
      OPP_ABBREVIATION = str_sub(MATCHUP, -3, -1)
    ) %>%
    left_join(
      abbrev_lookup,
      by = "OPP_ABBREVIATION"
    )
  
  # Get opponent win percentage table from df_teams
  opponent_win_pct <- df_teams %>%
    select(
      TEAM_NAME,
      W_PCT
    ) %>%
    rename(
      OPP_TEAM_NAME = TEAM_NAME,
      OPP_W_PCT = W_PCT
    )
  
  # Join opponent win percentage using opponent team name
  df_logs <- df_logs %>%
    left_join(
      opponent_win_pct,
      by = "OPP_TEAM_NAME"
    )
  
  # Average opponent win percentage for each team
  sos <- df_logs %>%
    group_by(
      TEAM_NAME
    ) %>%
    summarize(
      SOS_REG = mean(OPP_W_PCT, na.rm = TRUE),
      .groups = "drop"
    )
  
  return(sos)
}

head(get_sos("2015-16"))
```



4. This creates the main team-level dataset for one season and merges them with the playoff win percentage. 
```{python}
import time
import pandas as pd
from nba_api.stats.endpoints import leaguedashteamstats

def get_season_data(season_year):
    
    reg_adv = leaguedashteamstats.LeagueDashTeamStats(
        season = season_year,
        season_type_all_star = "Regular Season",
        measure_type_detailed_defense = "Advanced"
    )
    
    df_reg_adv = reg_adv.get_data_frames()[0]
    
    df_reg_adv = df_reg_adv[[
        "TEAM_ID",
        "TEAM_NAME",
        "W_PCT",
        "OFF_RATING",
        "DEF_RATING",
        "NET_RATING",
        "TS_PCT",
        "EFG_PCT",
        "OREB_PCT",
        "TM_TOV_PCT"
    ]]
    
    df_reg_adv = df_reg_adv.rename(columns = {
        "W_PCT": "W_PCT_REG",
        "OFF_RATING": "OFF_RATING_REG",
        "DEF_RATING": "DEF_RATING_REG",
        "NET_RATING": "NET_RATING_REG",
        "TS_PCT": "TS_PCT_REG",
        "EFG_PCT": "EFG_PCT_REG",
        "OREB_PCT": "OREB_PCT_REG",
        "TM_TOV_PCT": "TOV_PCT_REG"
    })
    
    df_reg_adv["PLUS_MINUS_REG"] = (
        df_reg_adv["OFF_RATING_REG"] - df_reg_adv["DEF_RATING_REG"]
    )
    
    time.sleep(1)
    
    reg_traditional = leaguedashteamstats.LeagueDashTeamStats(
        season = season_year,
        season_type_all_star = "Regular Season"
    )
    
    df_reg_traditional = reg_traditional.get_data_frames()[0]
    
    df_reg_traditional["THREE_PT_ATTEMPT_RATE_REG"] = (
        df_reg_traditional["FG3A"] / df_reg_traditional["FGA"]
    )
    
    df_reg_traditional = df_reg_traditional[[
        "TEAM_ID",
        "THREE_PT_ATTEMPT_RATE_REG"
    ]]
    
    time.sleep(1)
    
    po_traditional = leaguedashteamstats.LeagueDashTeamStats(
        season = season_year,
        season_type_all_star = "Playoffs"
    )
    
    df_po_traditional = po_traditional.get_data_frames()[0]
    
    df_po_traditional = df_po_traditional[[
        "TEAM_ID",
        "W_PCT"
    ]]
    
    df_po_traditional = df_po_traditional.rename(columns = {
        "W_PCT": "W_PCT_PO"
    })
    
    df = pd.merge(df_reg_adv, df_reg_traditional, on = "TEAM_ID")
    df = pd.merge(df, df_po_traditional, on = "TEAM_ID")
    
    df["season"] = season_year
    
    return df
get_season_data("2015-16").head()
```
5. Applies get_season_data to past 10 seasons.
```{python}
years = [
    "2015-16", "2016-17", "2017-18", "2018-19", "2019-20",
    "2020-21", "2021-22", "2022-23", "2023-24", "2024-25", "2025-26"
]

all_season_data = []

for year in years:
      season_df = get_season_data(year)
      all_season_data.append(season_df)

combined_df = pd.concat(all_season_data, ignore_index = True)

```
```{r}
head(py$combined_df)
```

6. Joins SOS data and creates SRS_REG to join with the main df. 
```{r}
combined_df <- py$combined_df

years <- c(
  "2015-16", "2016-17", "2017-18", "2018-19", "2019-20",
  "2020-21", "2021-22", "2022-23", "2023-24", "2024-25", "2025-26"
)

all_sos <- map_dfr(years, function(year) {
  Sys.sleep(5)
  get_sos(year) %>%
    mutate(season = year)
})

combined_df <- combined_df %>%
  left_join(
    all_sos,
    by = c("TEAM_NAME", "season")
  ) %>%
  mutate(
    SRS_REG = PLUS_MINUS_REG + SOS_REG
  )

head(combined_df)
```


7. Correlation per season with postseason win percentage
```{r}
cors <- combined_df %>% filter(season != "2025-26") %>%
  group_by(season) %>%
  summarize(
    win_pct_cor = cor(W_PCT_REG, W_PCT_PO),
    net_rating_cor = cor(NET_RATING_REG, W_PCT_PO),
    plus_minus_cor = cor(PLUS_MINUS_REG, W_PCT_PO),
    off_rating_cor = cor(OFF_RATING_REG, W_PCT_PO),
    def_rating_cor = cor(DEF_RATING_REG, W_PCT_PO),
    ts_pct_cor = cor(TS_PCT_REG, W_PCT_PO),
    efg_pct_cor = cor(EFG_PCT_REG, W_PCT_PO),
    three_pt_attempt_rate_cor = cor(THREE_PT_ATTEMPT_RATE_REG, W_PCT_PO),
    tov_pct_cor = cor(TOV_PCT_REG, W_PCT_PO),
    oreb_pct_cor = cor(OREB_PCT_REG, W_PCT_PO),
    sos_cor = cor(SOS_REG, W_PCT_PO),
    srs_cor = cor(SRS_REG, W_PCT_PO),
    .groups = "drop"
  )

cors
```
8. Average correalation over past ten seasons
```{r}
avg_cors <- cors %>%
  summarize(across(everything(), mean, na.rm = TRUE))

avg_cors
```
9. Correlations of variables with each other through matrix and colored matrix
```{r}
predictors <- combined_df %>% filter(season != "2025-26") %>%
  select(
    W_PCT_REG,
    NET_RATING_REG,
    PLUS_MINUS_REG,
    OFF_RATING_REG,
    DEF_RATING_REG,
    TS_PCT_REG,
    EFG_PCT_REG,
    THREE_PT_ATTEMPT_RATE_REG,
    TOV_PCT_REG,
    OREB_PCT_REG,
    SOS_REG,
    SRS_REG
  )

round(cor(predictors, use = "complete.obs"), 2)
```
```{r}
library(corrplot)

corrplot(
  cor(predictors, use = "complete.obs"),
  method = "color",
  type = "lower",
  addCoef.col = "black",
  number.cex = 0.4,
  tl.cex = 0.4
)
```
```{r}
write.csv(combined_df, "combined_df.csv", row.names = FALSE)
```

Based on the correlation with win percentage and  
1. NET_RATING_REG
2. TS_PCT_REG
3. THREE_PT_ATTEMPT_RATE_REG

## TO DO: 

1. Add more visuals
2. See how the difference in these variables correlates with the winner of each series

## NEXT DOCS

1) Data for matchups only
 1. how each team performed during regular season against team they are playing against and see how that correlates with the winner of the series
 2. Will want to add EDA visuals 
 
2) Modelling
 1. Choose 1 or 2 types of models (maybe Logistic Regression and RF), finalize our predictors (our target variable will be whether team A or team B wins)
 2. We can probably refer to PSTAT 131 Workflow and just copy that 

