---
title: "RF Model Playoff Series Prediction"
format: html
editor: visual
---

## RF Model

```{r, message=FALSE, warning=FALSE}
library(tidymodels)
library(ranger)
library(vip)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)

# -----------------------------
# Load Data
# -----------------------------
df <- read.csv("final_series_df.csv")

abbr_map <- c(
  "Atlanta Hawks" = "ATL",
  "Boston Celtics" = "BOS",
  "Brooklyn Nets" = "BKN",
  "Charlotte Hornets" = "CHA",
  "Chicago Bulls" = "CHI",
  "Cleveland Cavaliers" = "CLE",
  "Dallas Mavericks" = "DAL",
  "Denver Nuggets" = "DEN",
  "Detroit Pistons" = "DET",
  "Golden State Warriors" = "GSW",
  "Houston Rockets" = "HOU",
  "Indiana Pacers" = "IND",
  "LA Clippers" = "LAC",
  "Los Angeles Clippers" = "LAC",
  "Los Angeles Lakers" = "LAL",
  "Memphis Grizzlies" = "MEM",
  "Miami Heat" = "MIA",
  "Milwaukee Bucks" = "MIL",
  "Minnesota Timberwolves" = "MIN",
  "New Orleans Pelicans" = "NOP",
  "New York Knicks" = "NYK",
  "Oklahoma City Thunder" = "OKC",
  "Orlando Magic" = "ORL",
  "Philadelphia 76ers" = "PHI",
  "Phoenix Suns" = "PHX",
  "Portland Trail Blazers" = "POR",
  "Sacramento Kings" = "SAC",
  "San Antonio Spurs" = "SAS",
  "Toronto Raptors" = "TOR",
  "Utah Jazz" = "UTA",
  "Washington Wizards" = "WAS"
)

df <- df %>%
  mutate(
    TEAM_A_ABBR = abbr_map[TEAM_A],
    TEAM_B_ABBR = abbr_map[TEAM_B]
  )

df <- df %>%
  mutate(
    WINNER = factor(
      WINNER,
      levels = c(0, 1),
      labels = c("TEAM_B", "TEAM_A")
    )
  )

finals_obs <- df %>%
  filter(SERIES_GAMES == 1) %>%
  mutate(WINNER = NA,
         TEAM_A_ABBR = abbr_map[TEAM_A],
    TEAM_B_ABBR = abbr_map[TEAM_B])

df <- df %>%
  filter(SERIES_GAMES != 1)

# -----------------------------
# Train/Test Split
# -----------------------------
train_df <- df %>%
  filter(season != "2025-26")

test_df <- df %>%
  filter(season == "2025-26")

# -----------------------------
# Cross Validation
# -----------------------------
set.seed(31)

nba_folds <- vfold_cv(
  train_df,
  v = 5,
  strata = WINNER
)

# -----------------------------
# Recipe
# -----------------------------
nba_recipe <- recipe(
  WINNER ~
    WIN_PCT_DIFF +
    NET_RATING_DIFF +
    TS_DIFF +
    THREE_PT_DIFF +
    OFF_RATING_DIFF +
    DEF_RATING_DIFF +
    EFG_DIFF +
    OREB_DIFF +
    TOV_DIFF +
    PLUS_MINUS_DIFF +
    SRS_DIFF +
    H2H_TS_PCT_DIFF +
    H2H_WIN_PCT_DIFF +
    H2H_EFG_PCT_DIFF +
    H2H_POINT_DIFF_DIFF,
  data = train_df
)

# -----------------------------
# Random Forest Specification
# -----------------------------
rf_spec <- rand_forest(
  trees = 500,
  mtry = tune(),
  min_n = tune()
) %>%
  set_engine(
    "ranger",
    importance = "impurity"
  ) %>%
  set_mode("classification")

# -----------------------------
# Workflow
# -----------------------------
rf_workflow <- workflow() %>%
  add_recipe(nba_recipe) %>%
  add_model(rf_spec)

# -----------------------------
# Hyperparameter Grid
# -----------------------------
rf_grid <- expand.grid(
  mtry = c(2, 4, 6, 8, 10, 12, 15),
  min_n = c(2, 5, 10, 15, 20)
)

# -----------------------------
# Tune Model
# -----------------------------
set.seed(31)

rf_res <- tune_grid(
  rf_workflow,
  resamples = nba_folds,
  grid = rf_grid,
  metrics = metric_set(
    accuracy,
    roc_auc
  )
)

# -----------------------------
# View Results
# -----------------------------
collect_metrics(rf_res)
```
This table displays the performance metrics evaluated across the 5-fold cross-validation grid search. The grid covers combinations of two main hyperparameters: mtry (number of predictors randomly sampled at each split) and min_n (minimum number of data points required in a node to split it further)
```{r}
show_best(
  rf_res,
  metric = "roc_auc"
)
```
This table acts as a filter on the previous grid search, isolating the top 5 hyperparameter combinations that achieved the highest Area Under the Receiver Operating Characteristic curve (roc_auc)
```{r}

# -----------------------------
# Select Best Parameters
# -----------------------------
best_rf <- select_best(
  rf_res,
  metric = "roc_auc"
)

print(best_rf)
```
Best model had hyperparameters mtry = 2 and min_n = 20.
```{r}
# -----------------------------
# Finalize Workflow
# -----------------------------
final_rf <- finalize_workflow(
  rf_workflow,
  best_rf
)

# -----------------------------
# Fit Final Model
# -----------------------------
final_rf_fit <- fit(
  final_rf,
  data = train_df
)

# -----------------------------
# Predict on Test Set
# -----------------------------
rf_preds <- predict(
  final_rf_fit,
  test_df,
  type = "prob"
) %>%
  bind_cols(
    predict(final_rf_fit, test_df)
  ) %>%
  bind_cols(test_df)

# -----------------------------
# Accuracy
# -----------------------------
rf_preds %>%
  metrics(
    truth = WINNER,
    estimate = .pred_class
  )
```
This table shows the accuracy of the selected model on the test data.
```{r}

# -----------------------------
# ROC AUC
# -----------------------------
rf_preds %>%
  roc_auc(
    truth = WINNER,
    .pred_TEAM_A,
    event_level = "second"
  )
```
This table shows the AUC of the selected model on the test data.
```{r}

# -----------------------------
# Confusion Matrix
# -----------------------------
rf_preds %>%
  conf_mat(
    truth = WINNER,
    estimate = .pred_class
  )
```
This table breaks down where the model correctly and incorrectly predicted series results.
```{r}

# -----------------------------
# Feature Importance Plot
# -----------------------------
rf_imp <- final_rf_fit %>%
  extract_fit_parsnip() %>%
  vi()

rf_imp_top10 <- rf_imp %>%
  arrange(desc(Importance)) %>%
  slice_head(n = 10)

ggplot(rf_imp_top10, aes(
  x = reorder(Variable, Importance),
  y = Importance
)) +
  geom_col(fill = "#2C7FB8") +
  coord_flip() +
  labs(
    title = "Top 10 Most Important Predictors in NBA Playoffs",
    subtitle = "Random Forest (Impurity-based importance from ranger)",
    x = NULL,
    y = "Variable Importance"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40"),
    panel.grid.major.y = element_blank()
  )
```

```{r}

# -----------------------------
# Display Predictions
# -----------------------------
rf_preds_table <- rf_preds %>%
  mutate(
    Actual_Winner = if_else(WINNER == "TEAM_A", TEAM_A_ABBR, TEAM_B_ABBR),
    Predicted_Winner = if_else(.pred_class == "TEAM_A", TEAM_A_ABBR, TEAM_B_ABBR),
    Confidence = if_else(.pred_class == "TEAM_A", .pred_TEAM_A, .pred_TEAM_B),
    Correct = Actual_Winner == Predicted_Winner
  ) %>%
  select(
    TEAM_A = TEAM_A_ABBR,
    TEAM_B = TEAM_B_ABBR,
    Actual_Winner,
    Predicted_Winner,
    Confidence,
    Prob_TEAM_A = .pred_TEAM_A,
    Prob_TEAM_B = .pred_TEAM_B,
    Outcome = SERIES_SCORE,
    Correct
  ) %>%
  arrange(desc(Confidence))

rf_preds_table %>%
  mutate(Correct = NULL) %>%  # hide flag column in display
  kable(
    digits = 3,
    caption = "NBA Playoff Predictions Sorted by Confidence (Green = Correct)"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 12
  ) %>%
  row_spec(
    which(rf_preds_table$Correct),
    background = "#d4f7d4"  # light green
  )

finals_probs <- predict(final_rf_fit, finals_obs, type = "prob")
finals_class <- predict(final_rf_fit, finals_obs)

finals_prediction_table <- finals_obs %>%
  select(TEAM_A = TEAM_A_ABBR, TEAM_B = TEAM_B_ABBR) %>%
  bind_cols(finals_class, finals_probs) %>%
  mutate(
    Predicted_Winner = if_else(.pred_class == "TEAM_A", TEAM_A, TEAM_B),
    Confidence = if_else(.pred_class == "TEAM_A", .pred_TEAM_A, .pred_TEAM_B)
  ) %>%
  select(
    TEAM_A, 
    TEAM_B, 
    Predicted_Winner, 
    Confidence, 
    Prob_TEAM_A = .pred_TEAM_A, 
    Prob_TEAM_B = .pred_TEAM_B
  )

finals_prediction_table %>%
  kable(
    digits = 3,
    caption = "2026 NBA Finals: Random Forest Model Prediction"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 13
  ) %>%
  row_spec(
    1, 
    bold = TRUE, 
    background = "#FFF2CC" 
  )

# -----------------------------
# Save Predictions
# -----------------------------
write.csv(
  rf_preds_table,
  "rf_predictions_2025_26.csv",
  row.names = FALSE
)


```
